---
title: 'Modeling dynamics of Hansen & Hubbell 1980'
author: "Gaurav Kandlikar"
date: \today
output: pdf_document
---

I've written the following function to solve the differential equations. It uses the `R` package `deSolve`, which solves for the equations using Euler's methods. 
```{r, message=FALSE}
library(deSolve)
hh <- function(time,init,params) {
  with (as.list(c(time,init,params)), {
    
    # description of variables:
    # S0, D, mu1, y, Ks1, mu2, y2, KS2 are the params
    # S, N1, N2 are state variables that need to be in init
    
    dSdt  <- (S0-S)*D - (mu1/y1)*(S*N1)/(Ks1+S) - (mu2/y2)*(S*N2)/(Ks2+S)
    dN1dt <- (mu1*S*N1)/(Ks1+S) - (D*N1)
    dN2dt <- (mu2*S*N2)/(Ks2+S) - (D*N2)
    
    return(list(c(dSdt, dN1dt, dN2dt)))
  })
}
```

We can run the simulation usiung parameter values from Experiment 1. The expectation is that in the long run, the C-8 bacteria outcompetes PAO283, because both have identical $r$s but the former has a lower $K$.  

```{r experiment-1}
# Testing it out with parameters from Experiment 1
params1 <- c(S0 = 1e-4, D = 6e-2, 
            mu1 = .81, mu2 = .91, 
            y1 = 2.5e10, y2 = 3.8e10, 
            Ks1 = 3.0e-6, Ks2 = 3.1e-4)
inits  <- c(S = 1e-4, N1 = 100, N2 = 100000)

# Confirm that Js are distinct:
J1 <- with(as.list(params1), {(Ks1*(D/(mu1-D)))}); J1
J2 <- with(as.list(params1), {(Ks2*(D/(mu2-D)))}); J2

ex1 <- as.data.frame(ode(func = hh, y = inits, parms = params1, times = 1:100))
      
plot(log10(ex1$N2), type = "l", col = "darkred", lwd = 2, ylim = c(0, 8),
     main = "Dynamics predicted for Experiment 1", ylab = "log10 (Population size)", xlab = "time")
lines(log10(ex1$N1), col = "blue", lwd = 2)
legend("topright", lwd = 2, col = c("blue", "darkred"), 
       legend = c("C-8", "PAO283"), bty = "n")
```

Next, we run the simulation usiung parameter values from Experiment 2. The expectation is that in the long run, the C-8NalRSpecS outcompetes C-8 NalSSpecR, because both have identical $K$s but the former has a higher $r$.  
```{r experiment-2}
# Testing it out with parameters from Experiment 2
params2 <- c(S0 = 5e-6, D = 7.5e-2, 
            mu1 = .68, mu2 = .96, 
            y1 = 6.3e10, y2 = 6.2e10, 
            Ks1 = 1.6e-6, Ks2 = 1.6e-6)
inits  <- c(S = 5e-6, N1 = 10000, N2 = 1000)

# Confirm that Js are distinct:
J1 <- with(as.list(params2), {(Ks1*(D/(mu1-D)))}); J1
J2 <- with(as.list(params2), {(Ks2*(D/(mu2-D)))}); J2

ex2 <- as.data.frame(ode(func = hh, y = inits, parms = params2, times = 1:200))
      
plot(log10(ex2$N2), type = "l", col = "darkred", lwd = 2, ylim = c(0, 7),
     main = "Dynamics predicted for Experiment 2", ylab = "log(Population size)",
     xlab = "time")
lines(log10(ex2$N1), col = "blue", lwd = 2)
legend("topright", lwd = 2, col = c("blue", "darkred"), 
       legend = c("C-8 NalR SpecS", "C-8 NalS SpecR"), bty = "n")
```


Finally, we run the model with parameters from experiment 3. The expectation is that the two strains coexist indefinitely, as both have different values of $r$ and $K$ but in combination have similar values of $J$. 
```{r experiment-3}
# Testing it out with parameters from Experiment 3
params3 <- c(S0 = 5e-6, D = 7.5e-2, 
            mu1 = .68, mu2 = .41, 
            y1 = 6.3e10, y2 = 6.2e10, 
            Ks1 = 1.6e-6, Ks2 = 0.9e-6)
inits  <- c(S = 5e-6, N1 = 100, N2 = 100)

# Confirm that Js are nearly identical:
J1 <- with(as.list(params3), {(Ks1*(D/(mu1-D)))}); J1
J2 <- with(as.list(params3), {(Ks2*(D/(mu2-D)))}); J2

ex3 <- as.data.frame(ode(func = hh, y = inits, parms = params3, times = 1:100))
      
plot(log10(ex3$N2), type = "l", col = "darkred", lwd = 2, ylim = c(0, 7),
     main = "Dynamics predicted for Experiment 3", ylab = "log10 (Population size)", xlab = "time")
lines(log10(ex3$N1), col = "blue", lwd = 2)
legend("topright", lwd = 2, col = c("blue", "darkred"), 
       legend = c("C-8 NalR SpecS", "C-8 NalS SpecR"), bty = "n")
```

