---
title: 'EEB 200B: Lloyd-Smith PS 2'
author: "Gaurav Kandlikar"
date: \today
output: pdf_document
---
```{r echo = FALSE}
options(digits = 3)
```

### Question 1A: Complete the final two columns of the life table.  

|Age, $i$| Fecundity, $b_{i}$  | Number observed $n_{i}$ | Yearly survival, $s_{i}$ | Cumulative survival, $l_{i}$|
|:----:|:----:|:---------:|:---------:|:---------:|
|0|0.6|50|0.3| 1 |
|1|1.5|15|0  |0.3|
|2|0  |0 |-  |0  |

### Question 1B: Compute $R_{0}$ for this population. Do you expect it to grow or shrink over time? 
The basic reproductive rate $R_{0}$ can be calculated as
$$\sum_{n=1}^{i} l_{i}b_{i} = (1*0.6)+(0.3*1.5) = 1.05$$  
This means that each new offspring is expected to leave behind, on average, 1.05 offsprings of its own- so the population is expected to grow.

### Question 1C: Draw the life cycle graph for this population. How many age classes are needed?  

\begin{figure}[h]
  \centering
    \includegraphics[height=80mm]{figures/ps2_q1pc.pdf}
\end{figure}

### Question 1D: Write down the Leslie matrix that describes the dynamics of the system.  
$$\mathbf{L} = \left[\begin{array}
{rr}
0.6 & 1.5 \\
0.3 & 0
\end{array}\right]
$$

### Question 1E: Calculate the eigenvalues for this matrix by hand. Find the stable age distribution. 

I will calculate the eigenvalues by using the matrix determinant approach: 

$$\left[\bigg|\begin{array}
{rr}
0.6 & 1.5 \\
0.3 & 0
\end{array}\bigg| - \lambda \textbf{I} \right] = 0
$$

$$\left[\bigg|\begin{array}
{rr}
0.6 & 1.5 \\
0.3 & 0
\end{array}\bigg| - \lambda
\bigg|\begin{array}
{rr}
1 & 0 \\
0 & 1
\end{array}\bigg|\right] = 0
$$

$$\left[\bigg|\begin{array}
{rr}
0.6 & 1.5 \\
0.3 & 0
\end{array}\bigg| - 
\bigg|\begin{array}
{rr}
\lambda & 0 \\
0 & \lambda
\end{array}\bigg|\right] = 0
$$

$$\left[\bigg|\begin{array}
{rr}
(0.6-\lambda) & (1.5-0) \\
(0.3-0) & (0-\lambda)
\end{array}\bigg|\right] = 0
$$

$$(0.6 -\lambda)*(0-\lambda) - (0.3*1.5) = 0$$
$$\lambda^{2} - 0.6\lambda-0.45 = 0$$

Solving this quadratic equation for $\lambda$: 
```{r}
quad <- function(a,b,c) {
  to_return <- numeric(2)
  to_return[1] <- (-b+sqrt((b^2)-4*a*c))/(2*a)
  to_return[2] <- (-b-sqrt((b^2)-4*a*c))/(2*a)
  
  return(to_return)
}

lambdas <- quad(a = 1, b = -0.6, c = -0.45); lambdas
```

Here, $\lambda_{1}$ = `r lambdas[1]` and $\lambda_{2}$ = `r lambdas[2]`. The dominant eigenvalue `r lambdas[1]` is the long term population growth rate; the eigenvector associated with this eigenvalue will describe the stable age distribution.

To calculate the eigenvectors associated with $\lambda_{1}$, we can use the equation $\textbf{L}\bar{w}_{1} = \lambda_{1}\bar{w}_{1}$:

$$\bigg|\begin{array}
{rr}
0.6 & 1.5 \\
0.3 & 0
\end{array}\bigg| 
\bigg|\begin{array}
{r}
w_{11} \\
w_{12}
\end{array}\bigg| = \lambda 
\bigg|\begin{array}
{r}
w_{11} \\
w_{12}
\end{array}\bigg|
$$

$$0.6(w_{11})+1.5(w_{12}) = 1.03w_{11}$$
$$0.3(w_{11})+ 0 (w_{12}) = 1.03w_{12}$$
$$0.29(w_{11}) = w_{12}$$

```{r}
c12 <- .29/(1+.29)
c11 <- 1-c12
c(c11, c12)
```

### Question 1F: Calculate the eigenvalues and eigenvectors for this matrix with `R`.

```{r}
ll <- matrix(c(0.6, 1.5, 0.3, 0), byrow = T, ncol = 2); ll

ll_eigen <- eigen(ll)

# Take a look at the eigenvalues
ll_eigen$values
dom_eigval <- ll_eigen$values[1]
# Normalize the first eigenvector to get stable age distrib
sad <- as.matrix(ll_eigen$vectors[,1]/(sum(ll_eigen$vectors[,1]))); sad

```

### Question 1G: Confirm the result in part F by showing that the dominant eigenvector and eigenvalue satisfy the eigenvalue equation $L\bar{v}_{1} = \lambda_{i}\bar{v}_{1}$ 

```{r}
round(ll %*% sad, 5) == round(dom_eigval * sad, 5)
```

### Question 1H: A survey of variegated wombats across UCLA campus in 2015 observed population sizes of 2000 and 1000. What are the expected population sizes by age for 2016?  

```{r}
p2015 <- matrix(c(2000, 1000), ncol = 1)

p2016 <- ll %*% p2015; p2016
```

### Question 1I: If nothing else changes, by what proportion will population grow or shrink from the year 2067 to 2068? And from 2068 to 2069? 

The population should proportionally grow as $\lambda$- here, $\lambda$ = `r dom_eigval`

### Question 1J: The answer to part I is based on model prediction for long term growth of a population. Give three reasons why this prediction might not come true.  

Long term dynamics may be influenced by a number of parameters...

1. Variation in the environment may impact population sizes in a given year- an El Nino year, for example, might change the abundance of a food source- such a change in turn may drive $\lambda$ below 1 and drive the population to extinction! 

***More stuff!!***

### Question 1K: Use `R` to simulate the dynamics of the population for 50 time steps. Plot the absolute number in each age class and the proportion in each age class. Confirm predictions for long term population growth rate and stable age distrubution. 

```{r}
# First I will write a function to go through the generations 
project <- function(ll, init, t){
  to_return <- matrix(0, nrow = 2, ncol = t+1)
  to_return[, 1] <- init
  
  for(time in 2:(t+1)) {
    to_return[, time] <- as.vector(ll %*%(to_return[,time-1]))
  }
  return(to_return)
}

generations <- 50

population_trajectory <- project(ll, init = c(2000, 1000), generations)

prop <- function(x) {return(c(x[1]/sum(x), x[2]/sum(x)))}
population_trajectory_props <- apply(population_trajectory, 2, prop)

par(mfrow = c(1,2))
matplot(t(population_trajectory), ylab = "N", xlab = "Time", 
        main = "Population growth over time", lty = 1, lwd = 3, type = "l")
text(x = c(33, 40), y = c(10000, 1500), c("Age = 0", "Age = 1"))
matplot(t(population_trajectory_props), ylab = "N", xlab = "Time", xlim = c(-5, 50),
        main = "Population growth over time", lty = 1, lwd = 3, type = "l")
text(x = c(30, 30), y = c(0.7, 0.3), c("Age = 0", "Age = 1"))
abline(h = sad[1], col = "black", lty = 2, lwd = 2)
abline(h = sad[2], col = "red", lty = 2, lwd = 2)
```

### Question 1L: Will an absolute in crease of 0.01 in $b_{0}$ or $s_{0}$ have a greater impact on the long term growth rate of wombats? What about a relative increase of 1% in those two parameters? 

To get at this question, we need to calculate the Sensitivity and Elasticity matrices for this Leslie matrix. We can calculate the sensitivity matrix \textbf{S} as:
$$\mathbf{S} = \frac{\bar{v}_{1}\bar{w}_{1}^{T}}{\bar{v}_{1}\cdot\bar{w}_{1}}$$
Then, the elasticity matrix \textbf{E} can be calculated as
$$\mathbf{E} = \mathbf{S} \frac{L_{i,j}}{\lambda}$$

We still have not calculated the *left eigenvectors* for our Leslie Matrix. Let's do that first so that we have $v_{i}$s:

```{r}
left_eigens <- eigen(t(ll))
v1 <- left_eigens$vectors[,1]; v1

# let's also make a vector W11- currently the first eigenvector is stored as 'sad'
w1 <- sad
```

With both of these components in hand, we can use the formulae above to calculate \textbf{S} and \textbf{E}:

```{r}
# Sensitivity matrix:
ss <- (v1 %*% t(w1))/as.numeric(v1%*%w1)
print("The Sensitivty matrix:"); ss

ee <- ss*(ll/dom_eigval)
print("The Elasticity matrix:"); ee
```

The Sensitivity matrix \textbf{S} above predicts that an absolute increase of 0.01 in $S_{0}$ will have a greater impact on $\lambda$ than the same increase in $b_{0}$.

Conversely, the Elasticity matrix \textbf{E} predicts that a 1% increase in $b_{0}$ will impact $\lambda$ more than the same proportional increase in $s_{0}$. 

We can verify this prediction by manually perturbing the values of $s_{0}$ and $b_{0}$. First, let's verify for an absolue change of 0.01:
```{r}
# increase b0 by 0.01
ll_s1 <- matrix(c(0.6+0.01, 1.5, 0.3, 0), byrow = T, ncol = 2); ll_s1
lambda_s1 <- eigen(ll_s1)$values[1]

# increase s0 by 0.01
ll_s2 <- matrix(c(0.6, 1.5, 0.3+0.01, 0), byrow = T, ncol = 2); ll_s2
lambda_s2 <- eigen(ll_s2)$values[1]
```

Recall that above, we predicted that perturbing $s_{0}$ by 0.01 will have a greater impact on $\lambda$ than increasing $b_{0}$ by the same amount. Let's verify the prediction:
```{r}
lambda_s2 - dom_eigval > lambda_s1 - dom_eigval
```

Now, let's verify for a proportional increase of 1%:
```{r}
# increase b0 by 0.01
ll_e1 <- matrix(c(1.01*0.6, 1.5, 0.3, 0), byrow = T, ncol = 2); ll_e1
lambda_e1 <- eigen(ll_e1)$values[1]

# increase s0 by 0.01
ll_e2 <- matrix(c(0.6, 1.5, 1.01*0.3, 0), byrow = T, ncol = 2); ll_e2
lambda_e2 <- eigen(ll_e2)$values[1]
```

Recall that above, we predicted that perturbing $b_{0}$ by 1% will have a greater impact on $\lambda$ than increasing $s_{0}$ by the same proportion. Let's verify the prediction:
```{r}
lambda_e1 - dom_eigval > lambda_e2 - dom_eigval
```

Hurray!

### Question 1M: What is weird about the Elasticity matrix? Play around with different values in the Leslie matrix and develop an intuitive explanation. 

Let's recall the Elasticity matrix \textbf{E}:
```{r}
print("The elasticity matrix:"); ee
```

One surprising feature of this matrix is that the elasticities are equal for $b_{1}$ and $s_{0}$. The best intuitive explanation I can come up for this phenomenon is that both of these parameters eventually determine the reproductive output of a new-born individual one year from birth. So, the same proportional increase in either of these parameters will impact the reproductive value, and therefore $\lambda$ in the same way.  

It will be fun to iterate over a range of $s_{0}$s and check whether this holds:
```{r}
s0s <- seq(from = 0.5, to = 2.5, by = 0.1)
ratios_s0b1 <- numeric(length(s0s))
for(ii in 1:length(s0s)) {
  current_ll <- matrix(c(0.6, s0s[ii], 1.5, 0), byrow = T, ncol = 2)
  lambda <- eigen(current_ll)$values[1]
  w1 <- eigen(current_ll)$vectors[,1]
  v1 <- eigen(t(current_ll))$vectors[,1]
  ss <- (v1 %*% t(w1))/as.numeric(v1%*%w1)
  ee <- ss*(current_ll/lambda)
  ratios_s0b1[ii] <- ee[1,2]/ee[2,1]
}
ratios_s0b1
```

Sure enough, the ratio of the elasticies of $s_{0}$ and $b_{1}$ remains 1 for many values of $s_{0}$. To confirm, we can also iterate over a range of $b_{0}$s:
```{r}
b0s <- seq(from = 0.1, to = 1.5, by = 0.1)
ratios_s0b1 <- numeric(length(b0s))
for(ii in 1:length(b0s)) {
  current_ll <- matrix(c(b0s[ii], 0.3, 1.5, 0), byrow = T, ncol = 2)
  lambda <- eigen(current_ll)$values[1]
  w1 <- eigen(current_ll)$vectors[,1]
  v1 <- eigen(t(current_ll))$vectors[,1]
  ss <- (v1 %*% t(w1))/as.numeric(v1%*%w1)
  ee <- ss*(current_ll/lambda)
  ratios_s0b1[ii] <- ee[1,2]/ee[2,1]
}
ratios_s0b1
```

Woohoo! 

\clearpage 

## Question 2: Consider some more complex stage-structured populations. 

### Question 2A: Draw the life cycle graph corresponding to the Leslie matrix: 
$$\mathbf{L} = \left[\begin{array}
{rrrr}
0 & 0 & b_{s} & b_{A} \\
S_{Y} & 0 & 0 & 0 \\
0 & S_{J} & 0 & 0 \\
0 & 0 & S_{S} & S_{A} 
\end{array}\right]
$$


\begin{figure}[h]
  \centering
    \includegraphics[height=80mm]{figures/ps2_q2pa.pdf}
\end{figure}

### Question 2B: Write the Leslie matrix corresponding to this life cycle graph:  
\begin{figure}[h]
  \centering
    \includegraphics[height=80mm]{figures/ps2_q2pb.png}
\end{figure}
$$\mathbf{L} = \left[\begin{array}
{rrrrrrr}
0 & 0 & 0 & 0 & 0 & b_{5} & b_{6}  \\
p_{01} & 0 & 0 & 0 & 0 & 0 & c_{6} \\
p_{02} & p_{12} & 0 & 0 & 0 & 0 & 0\\
0 & p_{13} & p_{23} & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & p_{34} & 0  & 0 & 0 \\
0 & 0 & 0 & p_{35} & 0 & s_{5} & p_{65} \\
0 & 0 & 0 & p_{36} & p_{46} & p_{56} & s_{6} 
\end{array}\right]
$$

### Question 2C: Draw the life cycle graph corresponding to this Leslies matrix. What is a biological explanation for the stage classes labeled 3 and 4? 

$$\mathbf{L} = \left[\begin{array}
{rrrrr}
0 & b_{1} & b_{2} & b_{3} & 0 \\
s_{0} & 0 & 0 & 0 & 0 \\
0 & s_{1} & 0 & 0 & 0 \\
0 & 0 & s_{2} & s_{3}(1-p_{34}) & s_{4}p_{43}\\
0 & 0 & 0 & s_{3}p_{34} & s_{4}(1-p_{43})
\end{array}\right]
$$
\begin{figure}[h]
  \centering
    \includegraphics[height=80mm]{figures/ps2_q2pc.pdf}
\end{figure}

### Question 2D: Write an age structured Leslie matrix corrresponding to this life cycle. And write a stage-structured matrix with as few groups as possible. Is the stage structured matrix an exact match for this life cycle?
\begin{figure}[h]
  \centering
    \includegraphics[height=80mm]{figures/ps2_q2pd.png}
\end{figure}
$$\mathbf{L} = \left[\begin{array}
{rrrrrrr}
0 & 0 & b & b & b & b & b \\
S_{0} & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & S_{1} & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & S & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & S & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & S & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & S & 0 
\end{array}\right]
$$

The stage structured work can be done with a 3X3 Leslie matrix: 
$$\mathbf{L} = \left[\begin{array}
{rrrr}
0 & 0 & b & b \\
S_{0} & 0 & 0 & 0 \\ 
0 & S_{1} & 0 & 0 \\
0 & 0 & S & S
\end{array}\right]
$$

The full matrix tells us that this members of this species live for no longer than 6 years-  this information is lost when we make a Lefkovitch(!) matrix.
